Modeling and leveraging intuitive theories to improve vaccine attitudes

### 2022-09-30 paper/paper-papaja.Rmd
updated osf link
2nd resubmission

authornote: |
  Supplementary Materials, including all data and analysis code for this project are available at https://osf.io/sdfpj/. Preregistration information for Studies 3a and 3b is available at https://osf.io/csnez/registrations. Earlier modeling and analyses of data from Study 1 were presented at the 40th Annual Conference of the Cognitive Science Society. A six-page paper also appeared in the proceedings [@powell.etal2018].

@@

keywords          : "Intuitive theories; Belief revision; Vaccine attitudes; Cognitive modeling"
wordcount         : "9161"

@@

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over one thousand children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015; @lewandowsky.etal2012]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006]. This is especially  sobering in the context of the coronavirus pandemic: widespread reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed more than five million lives around the world.

@@

Other network-based approaches have also been applied to model systematic relationships among beliefs and attitudes, such as neural network models [e.g. see @holyoak.powell2016; @holyoak.simon1999; @simon.etal2004; @thagard1989; @spellman.etal1993]. In these models, beliefs or attitudes are represented by nodes (neurons) connected by undirected or bi-directed edges—in contrast to the directed edges used in Bayesian Networks. One application of neural-network models has been the modeling of “coherence-shifts”: patterns of belief revision wherein the beliefs or attitudes that should putatively serve as inputs to an evaluation are themselves modified during the process of evaluation. However, the directed edges of Bayesian networks are similarly capable of capturing these coherent reasoning dynamics: When conditioning on a node within a network, information flows in both directions along the edges. To illustrate, consider how observing smoke will lead a reasoner to infer that there is a fire, just as readily as observing fire will lead them to predict there will be smoke.[^1]

@@

To conduct an efficient search through the super-exponential space of graphs, we tested several learning algorithms implemented in the “bnlearn” R package [@scutari2010] and evaluated their performance through cross validation. To avoid overfitting, we used an 80% split of the data (*n* = 904) for training, and reserved the remaining 20% for evaluating model performance. We compared performance of different structure learning algorithms using k-fold cross validation within the training split. Graphs were scored by Bayesian Information Criteria, an asymptotic estimate of the likelihood of the data under a given graph structure. Based on these results, we chose to apply a score-based "hill-climbing"  algorithm using our generative model constraints (see Supplementary Materials for details). Applying these techniques resulted in an estimate of the network shown in Figure 1.

@@

Our goal in developing this cognitive model was to create a psychologically plausible and intelligible model of the intuitive theory surrounding vaccination decisions. However, it is also reasonable to ask whether other less-constrained models might be more predictive. To address this, we compared the Bayesian network model’s predictions to a suite of 14 naive “maximal” regression models, each using values on all 13 other belief scales as predictors for each belief. The model presented in Fig. 1---which we take to be simpler, more psychologically plausible, and more intelligible---performed as strongly as these "maximal" regression models (see Supplementary Materials).

@@

Very few educational interventions have been shown to improve attitudes toward childhood vaccination. One of the few successful educational interventions in this domain was tested by Horne and colleagues [-@horne.etal2015], who found that emphasizing the dangers of preventable diseases like mumps, measles, and rubella improved participants’ attitudes toward vaccines. Their intervention included descriptions of the symptoms of mumps, measles, and rubella; vivid photos of children suffering from these diseases; and a brief story from the perspective of a parent whose child was hospitalized with measles (see Supplementary Materials). This “disease risk” intervention improved attitudes toward vaccines compared to an irrelevant control essay and compared to an intervention aimed at debunking concerns about vaccines causing autism. In Study 2, we aimed to replicate this success and leverage the cognitive model developed in Study 1 to gain insight into how and why such an intervention successfully induces  belief revision in this domain.

>>>>

authornote: |
  Supplementary Materials, including all data and analysis code for this project are available at https://osf.io/sdfpj/. Preregistration information for Studies 3a and 3b is available at https://osf.io/uemq4. Earlier modeling and analyses of data from Study 1 were presented at the 40th Annual Conference of the Cognitive Science Society. A six-page paper also appeared in the proceedings [@powell.etal2018].

@@

keywords          : "Intuitive theories; Belief revision; Vaccine attitudes; Cognitive modeling"
wordcount         : "11853"

@@

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over one thousand children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015; @lewandowsky.etal2012]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006; and for some other effective approaches see e.g. @gagneur.etal2018; @opel.etal2013]. This is especially sobering in the context of the coronavirus pandemic: widespread reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed more than five million lives around the world.

@@

Other network-based approaches have also been applied to model systematic relationships among beliefs and attitudes, such as neural network models [e.g. see @holyoak.powell2016; @holyoak.simon1999; @simon.etal2004; @thagard1989; @spellman.etal1993]. In these models, beliefs or attitudes are represented by nodes (neurons) connected by undirected or bi-directed edges---in contrast to the directed edges used in Bayesian Networks. One application of neural-network models has been the modeling of “coherence-shifts”: patterns of belief revision wherein the beliefs or attitudes that should putatively serve as inputs to an evaluation are themselves modified during the process of evaluation. However, the directed edges of Bayesian networks are similarly capable of capturing these coherent reasoning dynamics: When conditioning on a node within a network, information flows in both directions along the edges. To illustrate, consider how observing smoke will lead a reasoner to infer that there is a fire, just as readily as observing fire will lead them to predict there will be smoke.[^1]

@@

To conduct an efficient search through the super-exponential space of graphs, we tested several learning algorithms implemented in the “bnlearn” R package [@scutari2010] and evaluated their performance through cross validation. To avoid overfitting, we used an 80% split of the data (*n* = 904) for training, and reserved the remaining 20% for evaluating model performance. We compared performance of different structure learning algorithms using k-fold cross validation within the training split. Graphs were scored by Bayesian Information Criteria, an asymptotic estimate of the likelihood of the data under a given graph structure. Based on these results, we chose to apply a score-based "hill-climbing"  algorithm using our generative model constraints (see Supplementary Materials for details). Applying these techniques resulted in an estimate of the network shown in Figure 2.

@@

Our goal in developing this cognitive model was to create a psychologically plausible and intelligible model of the intuitive theory surrounding vaccination decisions. However, it is also reasonable to ask whether other less-constrained models might be more predictive. To address this, we compared the Bayesian network model’s predictions to a suite of 14 naive “maximal” regression models, each using values on all 13 other belief scales as predictors for each belief. The model presented in Figure 2---which we take to be simpler, more psychologically plausible, and more intelligible---performed as strongly as these "maximal" regression models (see Supplementary Materials).

@@

Only a handful of educational interventions have been shown to improve attitudes toward childhood vaccination. One of the successful educational interventions in this domain was tested by Horne and colleagues [-@horne.etal2015], who found that emphasizing the dangers of preventable diseases like mumps, measles, and rubella improved participants’ attitudes toward vaccines. Their intervention included descriptions of the symptoms of mumps, measles, and rubella; vivid photos of children suffering from these diseases; and a brief story from the perspective of a parent whose child was hospitalized with measles (see Supplementary Materials). This “disease risk” intervention improved attitudes toward vaccines compared to an irrelevant control essay and compared to an intervention aimed at debunking concerns about vaccines causing autism. In Study 2, we aimed to replicate this success and leverage the cognitive model developed in Study 1 to gain insight into how and why such an intervention successfully induces  belief revision in this domain.
